Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-13 submission/manuscript.Rmd

- Motif analysis
	- only 1/3 had max/e2f - Unlikely to be the major recruitment mechanism for KDM5C
	- KDM5C does not contain motif-specific binding
	- controls cilia in mutiple cell types, initially identified in neurons. Shared axis that could contribute to NDDs
		- RFX Cilia in NDDs - https://pubmed.ncbi.nlm.nih.gov/33658631/ 
	- Enriched RFX transcription factors included the spermiogenesis regulator RFX2, whose mRNA is also significantly upregulated in _Kdm5c_-KO EpiLCs (Figure 4). 
	 - https://www.nature.com/articles/srep20435
	 - https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4498915/
	 - https://pubmed.ncbi.nlm.nih.gov/32725242/

- However, CPG islands greatly detrmine KDM5C recruitment - KDM5C is previously known to be enriched at CGIs. CGIs not typcially methylated, germline genes exception

We then globally characterized KDM5C binding at germline-enriched gene promoters through KDM5C ChIP-seq in EpiLCs and primary hippocampus and cortex neuron cultures (PNCs).

Our analysis of KDM5C ChIP-seq data provided novel insights into direct and indirect mechanisms by which germline genes can be misexpressed in _Kdm5c_-KO cells.

While we observed high enrichment of KDM5C peaks at germline genes in EpiLCs, KDM5C was not bound to germline gene promoters in PNCs. This suggests dysregulation of germline genes in the mature _Kdm5c_-KO brain is due to loss of repression during embryogenesis, which is consistent with previous work that found introducing human KDM5C into _Kdm5c_-KO PNCs does not repress two aberrant germline transcripts[@scandagliaLossKdm5cCauses2017]. Although enriched at germline gene promoters in EpiLCs, KDM5C was only bound to about a third of the germline genes aberrantly transcribed in _Kdm5c_-KO EpiLCs. One notable EpiLC DEG not bound by KDM5C is _Stra8_, a transcription factor activated by retinoic acid signaling in germ cells to promote meiotic initiation[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006]. However, retinoic acid can only activate _Stra8_ expression when DAZL is present. Unlike _Stra8_, _Dazl_ is a direct target of KDM5C in EpiCs and is both transcribed and translated in _Kdm5c_-KO EpiLCs. Expression of indirect DEGs like _Stra8_ indicates germline genes can be aberrantly expressed in _Kdm5c_-KO cells through downstream activation by other ectopic germline programs. These ectopic germline programs appear to loosely mimic the trajectory of typical germline development, as germline genes important for early germ cell development and meiosis are expressed in _Kdm5c_-KO EpiLC while late spermatogenesis genes are expressed in the mature _Kdm5c_-KO brain. Altogether, this indicates once a subset of germline genes are activated due to loss of direct KDM5C repression in the early embryo, these ectopic germline programs can continue to progress in the background of _Kdm5c_-KO somatic development.

@@

The above work provides the mechanistic foundation for KDM5C's role in germline gene repression and offers novel insight into how dysregulation in _Kdm5c_-KO tissue identity changes over the course of development. Although their impact upon neurodevelopment is currently unclear, ectopic germline transcripts are also found in models of other related neurodevelopmental disorders[@bonefasSomatogermlineTransformationChromatinlinked2021], including Immunodeficiency, Centromere instability, Facial anomalies syndrome 1 (ICF1 OMIM: #242860)[@velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], Kleefstra syndrome 1 (OMIM: #610253)[@schaeferControlCognitionAdaptive2009], and MeCP2 duplication syndrome (MDS, OMIM: #300260)[@samacoCrhOprm1Mediate2012]. Like KDM5C, the chromatin regulators underlying these conditions, DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2), primarily silence gene expression. Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that may have a similar underlying cause of germline versus soma dysregulation. However, further research is required to determine the ultimate impact of these germline genes and the extent to which this phenomenon occurs in humans.

- Last paragraph
	- Thus, failure to fine-tune the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_ could be one route by which loss of brain tissue identity contributes to _Kdm5c_-KO impairments.

@@

# Acknowledgements
- Jacob Mueller for providing insight in germline gene regulation. 
- Sundeep Kalantry for providing reagents and expertise in culturing mouse emrbyonic stem cells and epiblast-like cells
- Funding sources

>>>>

Our analysis of KDM5C ChIP-seq data provided novel insights into direct and indirect mechanisms by which germline genes can be misexpressed in _Kdm5c_-KO cells. While we observed high enrichment of KDM5C peaks at germline genes in EpiLCs, KDM5C was not bound to germline gene promoters in PNCs. This suggests dysregulation of germline genes in the mature _Kdm5c_-KO brain is due to loss of repression during embryogenesis, which is consistent with previous work that found introducing human KDM5C into _Kdm5c_-KO PNCs does not repress two aberrant germline transcripts[@scandagliaLossKdm5cCauses2017]. It is unclear how KDM5C is recruited to germline genes given that KDM5C itself does not contain domains for sequence-specific binding. In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018]. While MAX and E2F6 motifs were enriched at KDM5C-bound promoters in EpiLCs, only about one third of KDM5C-bound germline genes contained their consesus sequence at their promoter. Thus, other, unknown factors likely facilitate KDM5C's recruitment to germline genes.

We also found germline genes can become activated in _Kdm5c_-KO cells indepedent of direct loss of KDM5C suppression, given that KDM5C was only bound to about a third of germline-enriched DEGs. One notable EpiLC DEG not bound by KDM5C is _Stra8_, a transcription factor activated by retinoic acid signaling in germ cells to promote meiotic initiation[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006]. However, retinoic acid can only activate _Stra8_ expression when DAZL is present. Unlike _Stra8_, _Dazl_ is a direct target of KDM5C in EpiLCs and is transcribed _Kdm5c_-KO EpiLCs. Expression of indirect DEGs like _Stra8_ indicates germline genes can be aberrantly expressed in _Kdm5c_-KO cells through downstream activation by other ectopic germline programs. In support of this, germline genes unbound by KDM5C were significantly enriched for RFX transcription factors. RFX transcription factors bind to X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote the transcription of cilia and flagellar genes[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009]. One of the top RFX members enriched included RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016] whose mRNA is also expressed in _Kdm5c_-KO EpiLCs. Intruigingly, _Kdm5c_-KO EpiLCs primarily expressed germline genes involved in early germ cell development and meiosis while the mature _Kdm5c_-KO brain expressed late-stage spermatogenesis genes, including known RFX2 targets. Altogether, these data suggest once activated due to loss of direct KDM5C repression, ectopic germline programs can loosely mimic germline development and progress in the background of _Kdm5c_-KO somatic development.

@@

- CPG islands greatly detrmine KDM5C recruitment - KDM5C is previously known to be enriched at CGIs. CGIs not typcially methylated, germline genes exception. KDM5C promotes this process.

The above work provides the mechanistic foundation for KDM5C's role in germline gene repression and offers novel insight into how dysregulation in _Kdm5c_-KO tissue identity changes over the course of development. However, the contribution of ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown. KDM5C may be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes, such as _Apoc1_. KDM5C could also ensure tissue-specific timing of transcription factor family members, like RFX genes that are broadly required for cilia and flagella formation, including neuronal cilia[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009; @chungRFX2BroadlyRequired2012; @coschieraPrimaryCiliaPromote2024]. While their impact upon neurodevelopment is currently unclear, ectopic germline transcripts are also found in models of other related neurodevelopmental disorders[@bonefasSomatogermlineTransformationChromatinlinked2021], including Immunodeficiency, Centromere instability, Facial anomalies syndrome 1 (ICF1 OMIM: #242860)[@velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], Kleefstra syndrome 1 (OMIM: #610253)[@schaeferControlCognitionAdaptive2009], and MeCP2 duplication syndrome (MDS, OMIM: #300260)[@samacoCrhOprm1Mediate2012]. Like KDM5C, the chromatin regulators underlying these conditions, DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2), primarily silence gene expression. Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that have a similar underlying phenotype of germline versus soma dysregulation. Further research is required to determine the ultimate impact of these germline genes and the extent to which this phenomenon occurs in humans.

@@

# Acknowledgements
- Jacob Mueller for providing insight in germline gene regulation. 
- Sundeep Kalantry for providing reagents and expertise in culturing mouse emrbyonic stem cells and epiblast-like cells
- Ilakkiya
- Funding sources
